Transfer of Value Functions via Variational Methods

### 2018-05-18 NIPS 2018/main.tex
Edited sec 4

The second question that we need to answer is whether we can provide any guarantee on our algorithm's performance when given limited data. To address this point, we consider Algorithm~\ref{alg:transfer} with a mixture of Gaussians and linear approximators. We assume only a finite dataset is available and provide a finite-sample analysis bounding the distance between the fixed-point of the mellow Bellman operator and $Q$-functions sampled from the variational distribution minimizing the objective \eqref{eq:elbo}. Our main result is given in the following theorem.

\begin{restatable}{theorem}{thmg} \label{th:main-mixture}
Fix a target task $\tau$ and let $\wt{Q}$ be the fixed-point of the corresponding mellow Bellman operator. Assume linearly parameterized value functions $Q_{\bm{w}}(s,a)=\bm{w}^T\bm{\phi}(s,a)$ with uniformly bounded weights $|\bm{w}|\leq w_{max}$ and uniformly bounded features $|\bm{\phi}(s,a)|\leq \phi_{\max}$. Consider the mixture version of Algorithm~\ref{alg:transfer} using $C$ components, source task weights $\mathcal{W}_s$, and bandwidth $\sigma^2_p\ $ for the prior. Denote by $\wh{\bm{\xi}} = (\wh{\bm{\mu}}_1,\dots,\wh{\bm{\mu}}_C,\wh{\bm{\Sigma}}_1,\dots,\wh{\bm{\Sigma}}_C)$ the variational parameters minimizing the objective of Equation \eqref{eq:elbo} on a dataset $D$ of $N$ samples. Let $\nu$ be a probability measure over $\mathcal{S}\times\mathcal{A}$ and $\bm{w}^* = \arginf_{\bm{w}}\norm{\wt{B}_{\bm{w}}}_{\nu}^2$. Define $\upsilon(\bm{w}^*) \triangleq \mathbb{E}_{\mathcal{N}(\bm{w}^*,\bm{I})}\left[ v(\bm{w}) \right]$, with $v(\bm{w}) \triangleq \mathbb{E}_{\nu}\left[Var_{\mathcal{P}}\left[b(\bm{w})\right]\right]$. Then, there exist constants $c_1$, $c_2$, $c_3$, $c_4$ such that, with probability at least $1-2\delta$ over the choice of weights $\bm{w} \sim \frac{1}{C}\sum_i\mathcal{N}(\wh{\bm{\mu}}_i,\wh{\bm{\Sigma}}_i)$ and dataset $D$:
\begin{equation}
\resizebox{0.93\textwidth}{!}{$
\norm{Q_{\bm{w}} - \wt{Q}}_{\nu}^2 \leq \frac{1}{(1-\gamma)\delta}\left(2\norm{\wt{B}_{\bm{w}^{*}}}_{\nu}^2 + \frac{c_1}{N^2} + \frac{c_2 + \frac{c_4}{\abs{\mathcal{W}_s}}\sum_{j}\norm{\bm{w}^* - \bm{w}_{j}}}{N} + \frac{\upsilon(\bm{w}^*) + c_3\sqrt{\log\frac{2}{\delta}}}{\sqrt{N}}\right)$}.
\end{equation}
\end{restatable}

We refer the reader to App. \ref{} for the proof and a specific definition of the constants. Four main terms constitute our bound: the approximation error due to the limited hypothesis space (first term), the variance (second term), the distance to the prior (third term), and a constant term decaying as $\mathcal{O}(N^2)$. Intuitively, the bound is tighten when the source tasks' $Q$-functions are, on average, close to the optimal ones for the target task. In such case, the dominating error is due to the variance of the estimates, and, thus, the algorithm is expected to achieve good performance rather quickly, as new data is collected. Furthermore, as $N\rightarrow\infty$ the only error term remaining is, as usual, the irreducible approximation error due to the limited functional space. Notice that the variance term $\upsilon(\bm{w}^*)$ is due to the fact that we minimize a biased estimator of the Bellman error. If double sampling of the next state were possible (e.g., in simulation), such term could be removed. Finally, we want to point out that the proof of this theorem relies on a very general result (Lemma \ref{lemma:general-bound} in App. \ref{}) that does not require any specific choice of distribution of approximator. Leveraging on such results, it is straightforward to provide a finite-sample analysis of the Gaussian version of our algorithm. For the sake of completeness, we report the derivation in App. \ref{}.\todo{Should we discuss the main difference between the MGVT and GVT bounds here?}

@@

\begin{restatable}{theorem}{thmgsoft} \label{th:main-mixture-soft}
Fix a target task $\mathcal{M}_{\tau}$ and let $\wt{Q}$ be the fixed-point of the corresponding mellow Bellman operator. Assume linearly parameterized value functions $Q_{\bm{w}}(s,a)=\bm{w}^T\bm{\phi}(s,a)$ with bounded weights $\norm{\bm{w}}_2\leq w_{max}$ and uniformly bounded features $\norm{\bm{\phi}(s,a)}_2\leq \phi_{\max}$. Consider the mixture version of Algorithm~ \ref{alg:transfer} using $C$ components, source task weights $\mathcal{W}_s$, and bandwidth $\sigma^2_p\ $ for the prior. Denote by $\wh{\bm{\xi}} = (\wh{\bm{\mu}}_1,\dots,\wh{\bm{\mu}}_C,\wh{\bm{\Sigma}}_1,\dots,\wh{\bm{\Sigma}}_C)$ the variational parameters minimizing the objective of Eq. \eqref{eq:elbo} on a dataset $D$ of $N$ i.i.d. samples distributed according to $\tau$ and $\nu$. Let $\bm{w}^* = \arginf_{\bm{w}}\norm{\wt{B}_{\bm{w}}}_{\nu}^2$ and define $\upsilon(\bm{w}^*) \triangleq \mathbb{E}_{\mathcal{N}(\bm{w}^*,\frac{1}{N}\bm{I})}\left[ \text{v}(\bm{w}) \right]$, with $\text{v}(\bm{w}) \triangleq \mathbb{E}_{\nu}\left[Var_{\mathcal{P}_{\tau}}\left[\wt B_{\bm{w}}\right]\right]$. Then, there exist constants $c_1,c_2,c_3$ such that, with probability at least $1-2\delta$ over the choice of weights $\bm{w} \sim \frac{1}{C}\sum_i\mathcal{N}(\wh{\bm{\mu}}_i,\wh{\bm{\Sigma}}_i)$ and dataset $D$:
\begin{equation}
\resizebox{0.92\textwidth}{!}{$
\norm{Q_{\bm{w}} - \wt{Q}}_{\nu}^2 \leq \frac{1}{(1-\gamma)\delta}\left(2\norm{\wt{B}_{\bm{w}^{*}}}_{\nu}^2 + \upsilon(\bm{w}^*) + c_1\sqrt{\frac{\log\frac{2}{\delta}}{N}}  + \frac{c_2 +  \lambda d \log N + 2\lambda\varphi\left(\frac{1}{2\sigma^2_p}\norm{\bm{w}^* - \bm{w}_j}\right)}{N} + \frac{c_3}{N^2}\right)$},
\end{equation}
where, for a vector $\bm{x} = (x_1,\dots,x_d)$, $\varphi \left( x_j\right) \triangleq \sum_i \frac{e^{-x_i}}{\sum_j e^{-x_j}}x_i$ is the \textit{softmin} function.
\end{restatable}

>>>>

The second question that we need to answer is whether we can provide any guarantee on our algorithm's performance when given limited data. To address this point, we consider the two versions Algorithm~\ref{alg:transfer} of Sec. \ref{sec:gvt} and \ref{sec:mgvt} with linear approximators. We assume only a finite dataset is available and provide a finite-sample analysis bounding the distance between the fixed-point of the mellow Bellman operator and $Q$-functions sampled from the variational distribution minimizing the objective \eqref{eq:elbo}. Due to space constraints, we provide only the result for mixtures of Gaussians, while referring the reader to App. \ref{} for the Gaussian case.

\begin{restatable}{theorem}{thmg} \label{th:main-mixture}
Fix a target task $\mathcal{M}_{\tau}$ and let $\wt{Q}$ be the fixed-point of the corresponding mellow Bellman operator. Assume linearly parameterized value functions $Q_{\bm{w}}(s,a)=\bm{w}^T\bm{\phi}(s,a)$ with bounded weights $\norm{\bm{w}}_2\leq w_{max}$ and uniformly bounded features $\norm{\bm{\phi}(s,a)}_2\leq \phi_{\max}$. Consider the mixture version of Algorithm~ \ref{alg:transfer} using $C$ components, source task weights $\mathcal{W}_s$, and bandwidth $\sigma^2_p\ $ for the prior. Denote by $\wh{\bm{\xi}} = (\wh{\bm{\mu}}_1,\dots,\wh{\bm{\mu}}_C,\wh{\bm{\Sigma}}_1,\dots,\wh{\bm{\Sigma}}_C)$ the variational parameters minimizing the objective of Eq. \eqref{eq:elbo} on a dataset $D$ of $N$ i.i.d. samples distributed according to $\tau$ and $\nu$. Let $\bm{w}^* = \arginf_{\bm{w}}\norm{\wt{B}_{\bm{w}}}_{\nu}^2$ and define $\upsilon(\bm{w}^*) \triangleq \mathbb{E}_{\mathcal{N}(\bm{w}^*,\frac{1}{N}\bm{I})}\left[ \text{v}(\bm{w}) \right]$, with $\text{v}(\bm{w}) \triangleq \mathbb{E}_{\nu}\left[Var_{\mathcal{P}_{\tau}}\left[\wt B_{\bm{w}}\right]\right]$. Then, there exist constants $c_1,c_2,c_3$ such that, with probability at least $1-2\delta$ over the choice of weights $\bm{w} \sim \frac{1}{C}\sum_i\mathcal{N}(\wh{\bm{\mu}}_i,\wh{\bm{\Sigma}}_i)$ and dataset $D$:
\begin{equation}
\resizebox{0.92\textwidth}{!}{$
\norm{Q_{\bm{w}} - \wt{Q}}_{\nu}^2 \leq \frac{1}{(1-\gamma)\delta}\left(2\norm{\wt{B}_{\bm{w}^{*}}}_{\nu}^2 + \upsilon(\bm{w}^*) + c_1\sqrt{\frac{\log\frac{2}{\delta}}{N}}  + \frac{c_2 +  \lambda d \log N + 2\lambda\varphi\left(\frac{1}{2\sigma^2_p}\norm{\bm{w}^* - \bm{w}_j}\right)}{N} + \frac{c_3}{N^2}\right)$},
\end{equation}
where, for a vector $\bm{x} = (x_1,\dots,x_d)$, $\varphi \left( x_j\right) \triangleq \sum_i \frac{e^{-x_i}}{\sum_j e^{-x_j}}x_i$ is the \textit{softmin} function.
\end{restatable}

We refer the reader to App. \ref{} for the proof and a specific definition of the constants. Four main terms constitute our bound: the approximation error due to the limited hypothesis space (first term), the variance (second and third terms), the distance to the prior (third term), and a constant term decaying as $\mathcal{O}(N^2)$. Our main result shows a remarkable property of using Alg. \ref{alg:transfer} with mixtures of Gaussians: in order to tighten the bound, it is enough to have at least one source task that is close to the optimal solution of the target task. In such case, the dominating error is due to the variance of the estimates, and, thus, the algorithm is expected to achieve good performance rather quickly, as new data is collected. Furthermore, as $N\rightarrow\infty$ the only error terms remaining are the irreducible approximation error due to the limited functional space and the  variance term $\upsilon(\bm{w}^*)$. The latter is due to the fact that we minimize a biased estimator of the Bellman error and can be removed in cases where double sampling of the next state is possible (e.g., in simulation). Finally, we want to point out that the only difference between the bound of Thm. \ref{th:main-mixture} and the one for the Gaussian version of Alg. \ref{alg:transfer} (Thm. \ref{th:main-gaussian} in App. \ref{}) is, as expected, in the term bounding the distance to the prior. While in Thm. \ref{th:main-mixture} we have the (smoothened) minimum distance to the source tasks' weights, in the Gaussian case we have the distance to the mean of such weights, $\norm{\bm{w}^* - \bm{\mu}_p}_{\bm{\Sigma}_p^{-1}}$. This proves a remarkable advantage of using mixtures. In fact, the Gaussian version requires the source tasks to be, on average, similar to the target task in order to perform well, while the mixture version only requires this property for one of them. We verify this consideration from an empirical perspective in the next section.

@@

\thmg*
